Peer-to-Peer 6502 Cellular Automata

### 2018-07-03 paper.tex
removed time

% Macro to generate neighborhood memory map table
\newcommand\memtable[1]{
\begin{tabular}{lll}
  \hline
  From & To & Contents \\
  \hline
  #1
  \hline
\end{tabular}
}

@@

An {\bf update} of the CA consists of updating each cell in the CA in a pseudorandom order.
This can be any order that visits $L^3$ uniformly-random selected cells;
a Mersenne Twister is suitable for this,
and also for the generation of the random variable {\tt RUNTIME} used to constrain the update (see below).

@@

At the beginning of an update,
the registers {\tt A,X,Y} are initialized to zero,
all status flags (register {\tt P}) are set to zero,
the stack pointer (register {\tt SP}) is set to \hex{FF},
the program counter {\tt PC} is set to \hex{0000},
{\tt TIMELEFT} is initalized to {\tt STOPTIME} (see below),
and execution begins at the first byte of zero page.

Execution is terminated
when a software interrupt or undefined instruction is encountered,
or by non-maskable interrupt after a predetermined, but (pseudo)random, number of cycles {\tt STOPTIME}.
{\tt STOPTIME} is a 16-bit integer.
The LSB of {\tt STOPTIME} is uniformly distributed;
the MSB is distributed according to a geometric distribution with parameter $\frac{1}{2}$.
Thus, with probability $\frac{1}{2}$, the value of {\tt STOPTIME} lies uniformly in the range \hexrange{00}{FF};
with probability $\frac{1}{4}$, it lies in the range \hexrange{100}{1FF};
with probability $\frac{1}{8}$ it's in the range \hexrange{200}{2FF};
with probability $\frac{1}{16}$ it's \hexrange{300}{3FF};
and so on.
({\tt STOPTIME} stops incrementing and does not overflow when the MSB reaches \hex{FF},
which happens after the heat death of the universe.)
Throughout execution, the {\em remaining} value of this timer
(i.e. the number of cycles from now until termination)
is visible to the program
as the read-only 16-bit value called {\tt TIMELEFT},
available at memory address \hexrange{0400}{0401}.

@@

Setting the parameters $L$ and $C$ is a little arbitrary.
If we ran a 6502 CPU at 1Mhz for 10 minutes, with a mean of 400 cycles per cellular update
(the random variable {\tt STOPTIME} actually has a mean of 383.5 cycles),
we would update the whole board $10^6 \times 10 \times 60 \times 60 / (400 \times L^3)$ times
using the 10-minute bitcoin block rate.
The CA state requires $256 \times L^3$ bytes.
Faster clocks and protocols no doubt allow better numbers,
but considering these values and estimating conservatively,
we recommend $L=64$ and $C=256$ so that each block takes at most 64Mb to describe the CA state;
each cell is updated, on average, every few minutes;
and the proof-of-work burden (for $Z=1$) is roughly comparable to a 6502 CPU.

% Tables
\begin{table}
\begin{tabular}{llll}
  \hline
  From & To & Type & Contents \\
  \hline
  \hex{0000} & \hex{00FF} & RAM (storage-mapped) & State of cell $(x,y,z)$ \\
  \hex{0100} & \hex{01FF} & RAM (transient) & Stack, initialized to zero \\
  \hex{0200} & \hex{03FF} & RAM (transient) & Scratch memory, initialized to zero \\
  \hex{0400} & \hex{0401} & ROM (emulator-mapped) & {\tt TIMELEFT}: cycles left before termination \\
  \hex{0402} & \hex{3FFF} & Reserved & Read-only, zero \\
  \hex{4000} & \hex{7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
  \hex{8000} & \hex{FFFF} & ROM & Read-only (operating system) \\
  \hline
\end{tabular}
\caption{
  \label{tab:MemoryMap}
  Memory map for the 6502 virtual machine of cell $(x,y,z)$.
}
\end{table}

>>>>

@@

An {\bf update} of the CA consists of updating each cell in the CA in a pseudorandom order.
This can be any order that visits $L^3$ uniformly-random selected cells;
a Mersenne Twister is suitable for this.

@@

At the beginning of an update,
the registers {\tt A,X,Y} are initialized to zero,
all status flags (register {\tt P}) are set to zero,
the stack pointer (register {\tt SP}) is set to \hex{FF},
the program counter {\tt PC} is set to \hex{0000},
and execution begins at the first byte of zero page.

Execution is terminated
when a software interrupt or undefined instruction is encountered,
or by non-maskable interrupt after
completion of the first instruction that takes the cycle count past 65,536 cycles.

@@

Setting the parameters $L$ (CA size) and $C$ (updates per block)
is a little arbitrary, but can be guided by target memory footprint and clock rate.
For memory, the CA state requires $256 \times L^3$ bytes.
For clock rate, if we ran a 6502 CPU at 1Mhz for 10 minutes, with a mean of $2^{15}$ cycles per cellular update,
we would update the whole board $10^6 \times T / (2^{15} \times L^3)$ times
where $T$ is the time in seconds between blocks ($T=36,000$ for bitcoin).
Faster clocks and protocols no doubt allow better numbers,
but considering these values and estimating conservatively,
we recommend $L=64$ and $C=4$ so that
each cell is updated every few minutes;
each block takes at most 64Mb to describe the CA state;
and the computational burden is roughly comparable to a 6502 CPU.

% Tables, macros
% Macro to generate neighborhood memory map table
\newcommand\memtable[1]{
\begin{tabular}{lll}
  \hline
  From & To & Contents \\
  \hline
  #1
  \hline
\end{tabular}
}

% Tables
\begin{table}
\begin{tabular}{llll}
  \hline
  From & To & Type & Contents \\
  \hline
  \hex{0000} & \hex{00FF} & RAM (storage-mapped) & State of cell $(x,y,z)$ \\
  \hex{0100} & \hex{01FF} & RAM (transient) & Stack, initialized to zero \\
  \hex{0200} & \hex{03FF} & RAM (transient) & Scratch memory, initialized to zero \\
  \hex{0400} & \hex{3FFF} & ROM & Reserved, zero \\
  \hex{4000} & \hex{7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
  \hex{8000} & \hex{FFFF} & ROM & Operating system \\
  \hline
\end{tabular}
\caption{
  \label{tab:MemoryMap}
  Memory map for the 6502 virtual machine of cell $(x,y,z)$.
}
\end{table}
